Testing Gravity using Type Ia Supernovae Discovered by LSST

### 2018-12-12 doc/article.tex
before sending to Satya

\begin{abstract}
ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.2$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.  The increase in the number density of SNe~Ia, in parallel with improvements in the standardization of
their absolute magnitudes, now make them competitive probes of the growth of structure.  The peculiar velocity power spectrum
is sensitive to $f\sigma_8$,
the product of the linear growth and amplitude of density perturbations.  Cross-correlation with synergistic galaxy surveys further constrains $f \sigma_8$ and the galaxy bias. 
Thus, in the next decade the peculiar velocities of
SNe~Ia will provide the growth of structure  in the local $z<0.2$ Universe as a powerful test of General Relativity and other models of gravity.
\end{abstract}

@@

The probative power of a specific peculiar-velocity tracer primarily depends on its number density and the precision to which its absolute magnitudes are known.
The current generation of peculiar velocity studies use $10^3-10^5$ galaxies with Fundamental Plane and Tully-Fisher distances \citep{2008AJ....135.1738M, 2014MNRAS.445.2677S,
2016AJ....152...50T}. 
Next generation surveys WALLABY \citep{2008ExA....22..151J} and TAIPAN \citep{2017PASA...34...47D} are designed to increase these sample size by an order of magnitude.
These galaxies have absolute magnitude uncertainties of $\sim 0.4$~mag.

The current sample of SNe~Ia has a low number density compared to Fundamental Plane and Tully-Fisher galaxies.
Nevertheless, their low intrinsic-magnitude uncertainties can 
can provide peculiar velocities (expressed equivalently as peculiar magnitudes)
of their host galaxies \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}.  Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities that may be attributed to the growth of structure
\citet{2008MNRAS.389L..47A,2015JCAP...12..033H, 2017JCAP...05..015H}.
However, the ston is currently insufficient to perform a meaningful test of GR.

Two advances in the upcoming decade will make SNe~Ia  important probes of $f\sigma_8$.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\sigma_M \gtrsim 0.12$~mag absolute magnitude
dispersion.  However, SNe transmit more information than just the light curve shape and single color used in current analysis.
Recent studies indicate that with the right data, SNe absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.  This is a sample size comparable to those projected by WALLABY and TAIPAN.

@@

The projected precisions for LSST-discovered SNe~Ia have a number of interesting features.  Relative to
RSD, particular velocities contribute increasingly 
$f\sigma_8$ precision with decreasing redshift.  At redshift 0.3 the differential gain from adding peculiar velocities to the RSD measurement is small.
Despite the significant gain in volume and numbers of supernovae, the $f\sigma_8$ uncertainty in increasing redshift bins asymptotes such that
the differential gain in the cumulative  uncertainty is small for surveys with increasing maximum redshift.
For $z>0.1$, the volume is sufficiently large that we are not yet sample variance limited  given the LSST sample.

\section{Important Elements in Designing a Supernova Survey to Measure $\MakeLowercase{f}\sigma_8$}
\citet{2017ApJ...847..128H} perform their analysis in Fourier space, 
where the Fisher information matrix of a random Gaussian field with mean zero and covariance $C(k)$ parameterized by $\lambda$ is
approximated by
\begin{equation}
F_{ij} = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right],
\end{equation}
where the covariance for the velocity-velocity correlation
\begin{equation}
C = P_{vv}(k) + \frac{\sigma^2}{n}
\label{cov:eq}
\end{equation}
is dependent on the power spectrum, noise in the velocity measurement, and the density of velocity probes
\citep{2017MNRAS.464.2517H}.

>>>>

\begin{abstract}
ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.  The increase in the number density of SNe~Ia, in parallel with improvements in the standardization of
their absolute magnitudes, now make them competitive probes of the growth of structure.  The peculiar velocity power spectrum
is sensitive to $f\sigma_8$,
the product of the linear growth and amplitude of density perturbations.  Cross-correlation with synergistic galaxy surveys further constrains $f \sigma_8$ and the galaxy bias. 
Thus, in the next decade the peculiar velocities of
SNe~Ia will provide the growth of structure  in the local $z<0.3$ Universe as a powerful test of General Relativity and other models of gravity.
\end{abstract}

@@

The probative power of a specific peculiar-velocity tracer primarily depends on its number density and the precision to which its absolute magnitudes are known.
The current generation of peculiar velocity studies use $10^3-10^5$ galaxies with Fundamental Plane and Tully-Fisher distances \citep{2008AJ....135.1738M, 2014MNRAS.445.2677S,
2016AJ....152...50T}.  These galaxies have absolute magnitude uncertainties of $\sim 0.4$~mag. 
Next generation surveys WALLABY \citep{2008ExA....22..151J} and TAIPAN \citep{2017PASA...34...47D} are designed to increase these sample size by an order of magnitude
over half the sky to a depth of $z=0.1$.

The current sample of SNe~Ia has a low number density compared to Fundamental Plane and Tully-Fisher galaxies.
Nevertheless, their low intrinsic-magnitude uncertainties can 
can provide peculiar velocities (expressed equivalently as peculiar magnitudes)
of their host galaxies \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}.  Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities that may be attributed to the growth of structure
\citep{2008MNRAS.389L..47A,2015JCAP...12..033H, 2017JCAP...05..015H}.
However, the signal-to-noise is currently insufficient to perform a meaningful test of GR.

Two advances in the upcoming decade will make SNe~Ia  important probes of $f\sigma_8$.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\sigma_M \gtrsim 0.12$~mag absolute magnitude
dispersion.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SNe absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.  This is a sample size comparable to the number of galaxies projected by WALLABY and TAIPAN.

@@

The projected precisions for LSST-discovered SNe~Ia have a number of interesting features. 
Despite the significant gain in volume and numbers of supernovae, the $f\sigma_8$ uncertainty in increasing redshift bins asymptotes such that 
there is little benefit in going beyond
$z=0.3$.
The constraining power on  $f\sigma_8$ 
comes primarily from peculiar velocities
at $z<0.2$, but are dominated by RSD by $z=0.3$.
It turns out that for $z>0.1$, the volume is sufficiently large that the LSST sample is not yet sample variance limited.

\section{Important Elements in Designing a Supernova Survey to Measure $\MakeLowercase{f}\sigma_8$}
\citet{2017ApJ...847..128H} perform their analysis in Fourier space, 
where the Fisher information matrix of a random Gaussian field with mean zero and covariance $C(k)$ parameterized by $\lambda$ is
\begin{equation}
F_{ij} = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right],
\end{equation}
where the covariance for the velocity-velocity correlation
\begin{equation}
C = P_{vv}(k) + \frac{\sigma^2}{n}
\label{cov:eq}
\end{equation}
is dependent on the power spectrum, noise in the velocity measurement, and the density of velocity probes
\citep{2017MNRAS.464.2517H}.
